A Force-Balanced Control Volume Finite Element Method for Multiphase Porous Media Flow Modelling

### 2018-03-13 Papers/Prototype_Merged/article.tex

\documentclass[times]{fldauth}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\usepackage{graphicx}

\usepackage{epsfig}

@@

\usepackage{amssymb,amsmath,array}

\newcommand{\frc}{\displaystyle\frac}
\newcommand{\PN}[2][error]{P$_{#1}$DG-P$_{#2}$}
\newcommand{\PNDG}[2][error]{P$_{#1}$DG-P$_{#2}$DG}

@@

\address{\affilnum{1} Novel Reservoir Modelling and Simulation Group, Department of Earth Science and Engineering, Imperial College London, UK. \\ \affilnum{2}Applied Modelling and Computation Group, Department of Earth Science and Engineering, Imperial College London, UK. \\ \affilnum{3} Environmental and Industrial Fluid Mechanics Group, School of Engineering, University of Aberdeen, UK.}
\corraddr{Novel Reservoir Modelling and Simulation Group, Department of Earth Science and Engineering, Imperial College London, UK. E-mail: dimitrios.pavlidis@imperial.ac.uk}

\runningheads{J. L. M. A. Gomes \textit{et al.}}{A Force-Balanced CVFEM for Multi-phase Porous Media Flow Modelling}

\title{A Force-Balanced Control Volume Finite Element Method for Multi-phase Porous Media Flow Modelling}

\author{J. L. M. A. Gomes\affil{3,1}, D. Pavlidis\corrauth\affil{1,2}, P. Salinas\affil{1,2}, Z. Xie\affil{1,2}, J. R. Percival\affil{2}, \\Y. Melnikova\affil{1,2}, C. C. Pain\affil{1,2}, M. D. Jackson\affil{1}}

\begin{abstract}
  A novel method for simulating multi-phase flow in porous media is
  presented. The approach is based on a control volume finite element
  mixed formulation and new force-balanced finite element pairs. The
  novelty of the method lies in: (a) permitting both continuous and
  discontinuous description of pressure and saturation between
  elements; (b) the use of arbitrarily high-order polynomial
  representation for pressure and velocity and (c) the use of
  high-order flux-limited methods in space and time to avoid
  introducing non-physical oscillations while achieving high-order
  accuracy where and when possible. The model is initially validated
  for two-phase flow. Results are in good agreement with analytically
  obtained solutions and experimental results. The potential of this
  method is demonstrated by simulating flow in a realistic geometry
  composed of highly permeable meandering channels.
\end{abstract}

\keywords{CVFEM mixed formulation; Discontinuous Galerkin; Flux-limiting; High-order method; Multi-phase flows; Porous media flows}

\maketitle

\section{Introduction}

Numerical modelling of multi-phase flows in porous media has
applications in a wide range of disciplines, including hydrocarbon and
groundwater production, safety assessment of deep geological disposal
of radioactive waste, and carbon capture and storage \cite{chen_2006,
  aiea_1999, pruess_1990c, jiang_2011}. Such applications often
include complex geometries with irregular (and often internal)
boundaries between rock types with contrasting porosity and
permeability. Accurately capturing these complex geometries is a key
challenge when simulating such flows.

Finite difference methods (FDM) have been extensively used for
modelling fluid flows in porous media
\cite{aziz_1986, chen_1997, chen_2005}. However, they are strongly
dependent on mesh quality and orientation, and cannot easily represent
complex geometries. In addition, FDM can produce excessive numerical
dispersion in heterogeneous porous media \cite{chavent_1986}.

The geometrical flexibility of finite element methods (FEM) has been
shown to overcome these deficiencies. Among FEM-based formulations for
porous media, the control volume finite element method (CVFEM) is
increasingly popular as it can guarantee local mass conservation, it
has the potential to be high-order accurate, and is able to use
tetrahedral, geometry-conforming elements \cite{forsyth_1990,
  cordazzo_2004, geiger_2004, hurtado_2007}. Huber and Helmig \cite{huber_2000}
demonstrated that the vertex-centred, finite volume-finite element
method (or Box scheme) can achieve similar goals (see also \cite{helmig_1997}).

Durlofsky \cite{durlofsky_1993,durlofsky_1994} compared the performance of a
mixed FEM formulation (piece-wise linear in velocity and piece-wise
constant in pressure) and the classical CVFEM for single phase Darcy
flows. It was concluded that the latter is computationally more
efficient and numerically more accurate than the former for problems
involving a (relatively) small number of degrees of freedom. Moreover,
for complex heterogeneous problems with sharp changes in material
properties, the mixed FEM formulation often led to physically
unrealistic solutions.  Cumming et al. \cite{cumming_2011} demonstrated that a
CVFEM-based discretisation could be used to solve the Richards
equation (coupled mass conservation and Darcy equations) in
heterogeneous porous media with relatively small computational
overhead, compared with traditional, coupled velocity-pressure based
formulations.  Mass balance was enforced as described by Kirkland et al.
\cite{kirkland_1992} (see also \cite{forsyth_1990,cumming_phd2012}). However, CVFEM often requires
high resolution meshes in regions where material properties vary
abruptly, such as permeability contrasts at (for example)
fracture-matrix interfaces, or boundaries between different rock
types. Control volume (CV) boundaries span finite elements where
material properties are defined; therefore, some average value of the
permeability must be calculated across CV interfaces. This often leads
to excessive numerical dispersion, especially in highly heterogeneous
media \cite{nick_2011b, nick_2011a}.

Discontinuity-capturing schemes (e.g. shock waves, contact
  surface or material discontinuities --
  see \cite{brooks_1982,tezduyar_1986}) were originally developed to
resolve sharp changes in solution fields such as velocity or
saturation. Related discontinuity-capturing schemes include the
discontinuous-Galerkin FEM (DGFEM) scheme in which continuity of the
solution is not explicitly enforced, allowing sharp changes in the
solution fields to be captured. The DGFEM scheme is stabilised,
locally conservative and designed to achieve high-order
accuracy. DGFEM solution fields are allowed to be discontinuous at the
element faces; thus, the solution is able to handle discontinuities in
material properties at internal boundaries. In addition, DGFEM is
well-suited to deal with interface problems by incorporating
specially-designed interface fluxes. These properties have attracted
the attention of the porous media flow community over the past 15
years (see \cite{riviere_2000,riviere_2002,bastian_2002}) and are
utilised here.

In this paper, a novel CVFEM formulation which is conservative and
consistent is presented. The continuity equations are embedded into
the pressure equation to enforce mass conservation and the exact force
balance. The \PN[n]{m}[DG] family of triangular and tetrahedral finite
element pairs is used to discretise velocity and pressure in
space. For this element type, the velocity field is represented by
$n^{\rm th}$-order polynomials that are discontinuous across elements,
while the pressure field is represented by $m^{\rm th}$-order
polynomials that may be continuous (termed P$_n$DG-P$_m$) or
discontinuous (termed P$_n$DG-P$_m$DG) across elements. Control
volumes are dual to the pressure mesh. The formulation employs an
implicit algorithm with respect to time that is less restrictive than
the implicit-pressure-explicit-saturation (IMPES) scheme often adopted
in porous media flow problems (e.g. see \cite{aziz_1986, geiger_2004}).

The method is demonstrated using two families of element pairs: the
\PN[n]{m} pair which can allow the velocity to exactly represent the
pressure gradients in the flow solution for homogeneous material
properties; and the \PNDG[n+1]{n} pair, which has similar properties
to the \PN[n]{n+1} element pair, but allows a representation in which
pressure, saturation and other solution variables are discontinuous
across finite element boundaries. This solves the long-standing
problem described above with respect to the use of traditional CVFEM
to capture sharp changes in material properties.

Results using this method in simple geometries have been previously
reported in the literature
(e.g. see \cite{jackson_2015,kaisu}). However, the method has not
been described in detail or thoroughly validated, and its potential has not been
explored. The remainder of this paper is organised as follows. The
governing equations and numerical methods employed to solve them are
presented in Section~\ref{overlapping_method_section}. Model set-up
and results including comparison against reference data are given in
Sections~\ref{setup} and \ref{res}, respectively. Finally, some
concluding remarks are presented in Section~\ref{conc}.

\section{Governing equations and numerical method}
\label{overlapping_method_section}

\subsection{A novel representation of multi-phase Darcy flows}
Darcy's law for immiscible multi-phase flow may be written in the form:
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} =
  -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(
  \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity,
$\mathbf{K}$ is the absolute permeability tensor of the porous medium,
$\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase
relative permeability, which is a function of the phase saturation
$S_{\alpha}\left(\mathbf{r},t\right)$. $\mu_{\alpha}$, $p_{\alpha}$,
$\rho_{\alpha}$, and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic
viscosity, pressure, density and source term, which may include gravity and/or
capillarity, respectively.

@@

In order to discretise Eqn.~\ref{force-bal}, a finite element
representation for $\mathbf{v}_\alpha$ and $p$ is assumed, expressed
in terms of their FE basis functions $Q_{j}$ and $P_{j}$,
respectively, as:
\begin{equation}
  \mathbf{v}_\alpha(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_u}
  Q_{j}(\bm{r})\mathbf{v}_{\alpha,j}(t) \;\;\;\;\text{ and } \;\;\;\;
  p(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_p}
  P_{j}(\bm{r})p_{j}(t).
\end{equation} 
Here $\mathcal{N}_{u}$ and $\mathcal{N}_{p}$ are the total number of
degrees of freedom for the FE force and pressure representations. Each
component of the weak form of the force balance (Eqn.~\ref{force-bal})
is tested with the $\mathbf{v}_\alpha$ basis function space to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i \left (
             {\mathbf v}_\alpha + \nabla p  -\rho_\alpha g \mathbf{d} -{\mathbf s}_{u_\alpha}
             \right) dV \right. + \displaystyle \oint_{\Gamma_{E}}
                   {Q}_i {\mathbf n} \left(p - \tilde{p}\right)
                   d\Gamma + \nonumber \\ \oint_{\Gamma_{\Omega}} {
                     Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma
                   = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of
element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of
the computational domain. The numerical pressure $\tilde{p}$ appearing
in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is
the arithmetic mean of the potentially discontinuous pressure across
the element $E$. This term vanishes when a continuous formulation is
used to discretise the pressure field. The last term in
Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure
level to $p_\text{bc}$ on a computational domain boundary.

@@

\subsection{Control volume finite element types}
In the formulation presented here, pressure, velocity, permeability
and porosity are represented FE-wise, but saturation, relative
permeability and fluid properties such as viscosity and density are
represented CV-wise. Fig.~\ref{fem_cv_represent_a} displays the two
families of element types presented in this
paper. Fig.~\ref{fem_cv_represent_a} (a) shows the P$_{n}$DG-P$_{n+1}$
element type, with $n = 1$, in two dimensions (2D). Here, the velocity
field is linear, discontinuous (between elements), the pressure field
has a quadratic, continuous (between elements), representation and the
CV's span various elements. Fig.~\ref{fem_cv_represent_a} (b) shows
the P$_{n+1}$DG-P$_{n}$DG element type, with $n = 1$, also in 2D. Here,
the velocity field is quadratic, discontinuous (between elements), the
pressure field has a linear, discontinuous (between elements),
representation and the CV's do not span elements.

\begin{figure}[h!]
\subfloat[P$_1$DG-P$_2$]{\includegraphics[width=.5\textwidth]{p1dgp2-cont-sat.pdf}}
\subfloat[P$_2$DG-P$_1$DG]{\includegraphics[width=.5\textwidth]{p2dgp1-dg-sat.pdf}}
\caption{2D representation of the element pairs presented in this
  work. Shaded areas denote control volumes (in which saturation is
  stored), black points represent the pressure nodes and white points
  the velocity nodes.}
    \label{fem_cv_represent_a}
\end{figure}

@@

\subsection{Saturation and global mass conservative equations} \label{Section:Saturation_Global}
Here, the saturation conservation equations are discretised and the
global mass balance equation is derived. In this work, fluids are
assumed incompressible. The saturation equation can be written as:
\begin{equation}
  \phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} + \nabla
  \cdot \left( {\mathbf u}_{\alpha} S_{\alpha}\right) =
  s_{cty,\alpha},
  \label{saturation_equation}
\end{equation}
where $\phi$ is the porosity and $s_{cty}$ is a source
term. Eqn.~\ref{saturation_equation} is discretised in space by
testing with CV basis functions $M_{i}$ and with the $\theta$-method
in time:
\begin{eqnarray}
  \int_{\Omega} M_{i} \displaystyle\frac{\phi \left({S_{\alpha
        i}^{n+1}}-{S_{\alpha i}^{n}}\right)}{\Delta t} dV + \nonumber
  \\ \oint_{\Gamma_{CV_{i}}} \left[\theta^{n+\frac{1}{2}} {\mathbf n}
    \cdot {\mathbf u}_{\alpha}^{n+1} S_{\alpha}^{n+1} +
    \left(1-\theta^{n+\frac{1}{2}}\right) {\mathbf n} \cdot {\mathbf
      u}_{\alpha}^{n} S_{\alpha}^{n} \right]d\Gamma = \int_{\Omega}
  M_{i} {s_{cty,\alpha}^{n+\theta}} dV,
  \label{detail-sat-eqn-k}
\end{eqnarray}
where $\mathbf{n}$ is the outward-pointing unit normal vector to the
surface ($\Gamma_{CV_{i}}$) of the control volume $i$ and $n$ is the
current time level. More details on the numerical methods used to
solve Eqn.~\ref{detail-sat-eqn-k} can be found in
\cite{pavlidis_2013b}.

>>>>

%%
%% Copyright 2007, 2008, 2009 Elsevier Ltd
%%
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%%
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later c.
%%
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%%

%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references
%% SP 2008/03/01
%%
%%
%%
%% $Id: elsarticle-template-harv.tex 4 2009-10-24 08:22:58Z rishi $
%%
%%
\documentclass[preprint,authoryear,12pt]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,authoryear,1p,times]{elsarticle}
%% \documentclass[final,authoryear,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,authoryear,3p,times]{elsarticle}
%% \documentclass[final,authoryear,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,authoryear,5p,times]{elsarticle}
%% \documentclass[final,authoryear,5p,times,twocolumn]{elsarticle}

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%\usepackage{cases}
%% or use the graphicx package for more complicated commands
\usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
\usepackage{epsfig}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb,amsmath,array}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% natbib.sty is loaded by default. However, natbib options can be
%% provided with \biboptions{...} command. Following options are
%% valid:

%%   round  -  round parentheses are used (default)
%%   square -  square brackets are used   [option]
%%   curly  -  curly braces are used      {option}
%%   angle  -  angle brackets are used    <option>
%%   semicolon  -  multiple citations separated by semi-colon (default)
%%   colon  - same as semicolon, an earlier confusion
%%   comma  -  separated by comma
%%   authoryear - selects author-year citations (default)
%%   numbers-  selects numerical citations
%%   super  -  numerical citations as superscripts
%%   sort   -  sorts multiple citations according to order in ref. list
%%   sort&compress   -  like sort, but also compresses numerical citations
%%   compress - compresses without sorting
%%   longnamesfirst  -  makes first citation full author list
%%
%% \biboptions{longnamesfirst,comma}

% \biboptions{}

%\journal{Computer Methods in Applied Mechanics and Engineering}
\journal{Comput Meth Appl Mech Eng}
\newcommand{\frc}{\displaystyle\frac}
\newcommand{\PN}[2][error]{P$_{#1}$DG-P$_{#2}$}
\newcommand{\PNDG}[2][error]{P$_{#1}$DG-P$_{#2}$DG}

@@

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

\title{A Force-Balanced Control Volume Finite Element Method for Multiphase Porous Media Flow Modelling}

%% use optional labels to link authors explicitly to addresses:
\author[UoA,NORMS]{J. L. M. A. Gomes}\corref{cor1}\ead{jefferson.gomes@abdn.ac.uk} \author[IC,NORMS]{D. Pavlidis} 
\author[IC,NORMS]{P. Salinas} \author[IC,NORMS]{Z. Xie}
\author[IC]{J. R. Percival} \author[NSW]{P. Mostaghimi} 
\author[IC,NORMS]{C. C. Pain} \author[NORMS]{M. D. Jackson}
%\author[IC]{M. Blunt}

\cortext[cor1]{Corresponding author.}
\address[UoA]{Environmental and Industrial Fluid Mechanics Group, School of Engineering, University of Aberdeen, UK}
\address[IC]{Applied Modelling and Computation Group, Department of Earth Science and Engineering, Imperial College London, UK}
\address[NSW]{School of Petroleum Engineering, University of New South Wales, Australia}
\address[NORMS]{Novel Reservoir Modelling and Simulation Group, Department of Earth Science and Engineering, Imperial College London, UK}

\begin{abstract}
  A novel control volume finite element method for simulating
  subsurface multiphase flows is presented here. The approach has two
  major strengths. First, it uses a dual consistent pressure-velocity
  representation in control volume and finite element spaces which
  ensures local mass conservation. Second, it uses a family of
  finite element types that ensure the multiphase Darcy equations can be
  exactly enforced. The underlying mass conservation equations are
  solved in control volume space whereas finite elements are used to
  obtain high-order fluxes on control volume boundaries. The novelty
  of the method lies in (a) permitting both continuous and
  discontinuous description of pressures and saturations between
  elements; (b) the use of arbitrarily high-order polynomial
  representation for pressure and velocity within the mixed
  formulation and (c) the use of high-order flux-limited methods to
  avoid introducing non-physical oscillations while achieving
  high-order time accuracy where and when possible. The model is
  initially evaluated and then applied to a series of test cases using
  structured and unstructured (triangular/ tetrahedral) meshes, as
  well as heterogeneous problems with large permeability
  ratios. Numerical results are in good agreement with analytically
  obtained solutions.
\end{abstract}

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
CVFEM mixed formulation \sep Discontinuous Galerkin  \sep Multiphase flows \sep Porous media.

\end{keyword}

\end{frontmatter}

\section{Introduction}
Numerical modelling of multiphase flow in porous media has
applications in a wide range of disciplines, including hydrocarbon and
groundwater production, safety assessment of deep geological disposal
of radioactive waste, and CO$_{\text{2}}$ migration and trapping in
carbon capture and storage ~\citep{chen_2006, aiea_1999, pruess_1990c,
  jiang_2011}. Such geological and geophysical applications often
include complex rock geometries with irregular (and often internal)
boundaries between rock types with contrasting porosity and
permeability.  Accurately capturing these complex geometries is a key
challenge when simulating such flows.

Finite difference methods (FDM) have been extensively used in
modelling and simulation of fluid flows in porous media
\citep{aziz_1986, chen_1997, chen_2005}.  However, they are strongly
dependent on mesh quality and orientation, and cannot easily represent
complex geometries. In addition, FDM can produce excessive numerical
dispersion in heterogeneous porous media \citep{chavent_1986}. The
geometrical flexibility of finite element methods (FEM) has been shown
to overcome these deficiencies. Among FEM-based formulations for
porous media, the control volume finite element method (CVFEM) has
become increasingly popular as it can guarantee local mass
conservation and high-order numerical accuracy, as well as being able
to use tetrahedral, geometry-conforming elements \citep{forsyth_1990,
  cordazzo_2004, geiger_2004, hurtado_2007}. \citet{huber_2000}
demonstrated that the vertex-centred, finite volume-finite element
method (or Box scheme) can achieve similar goals \citep[see
  also][]{helmig_1997}.

\citet{durlofsky_1993,durlofsky_1994} compared the performance of a
mixed FEM formulation (i.e. piece-wise constant in pressure and
piece-wise linear in velocity) and the classical CVFEM for single
phase Darcy flows. He concluded that the latter is more
computationally efficient and numerically accurate than the former for
problems involving a (relatively) small number of degrees of
freedom. Moreover, for complex heterogeneous problems with sharp
changes in material properties, mixed FEM formulations often lead to
physically unrealistic solutions.  \citet{cumming_2011} demonstrated
that a CVFEM-based discretisation could be used to solve the Richards
equation (coupled mass conservation and Darcy equations) in
heterogeneous porous media with relatively small computational
overhead, compared with traditional, coupled velocity-pressure based
formulations.  Mass balance was enforced as described by
\citet{kirkland_1992}~\citep[see
  also][]{forsyth_1990,cumming_phd2012}.  The efficiency and accuracy
of CVFEM for porous media flows has motivated our development of a
CVFEM-based discretisation here.  However, CVFEM often requires high
resolution meshes in regions where material properties vary abruptly,
such as permeability contrasts at (for example) fracture-matrix
interfaces, or boundaries between different rock types. Control volume
(CV) boundaries span finite elements where material properties are
defined; therefore, some average value of the permeability must be
calculated across CV interfaces. This often leads to excessive
numerical dispersion, especially in highly heterogeneous media
\citep{nick_2011b, nick_2011a}.

Discontinuity-capturing schemes~\citep[e.g. shock waves, contact
  surface or material discontinuities --
  see][]{brooks_1982,tezduyar_1986} were originally developed to
resolve sharp changes in solution fields such as velocity or
saturation. Related discontinuity-capturing schemes include the
discontinuous-Galerkin FEM (DGFEM) scheme in which continuity of the
solution is not explicitly enforced, allowing sharp changes in the
solution fields to be captured. The DG formulation is a stabilised
finite element method that is locally conservative and designed to
achieve high-order accuracy. DGFEM solution fields are allowed to be
discontinuous at the element faces; thus, the solution is able to
handle discontinuities in material properties at internal
boundaries. In addition, DGFEM is well-suited to deal with interface
problems by incorporating specially-designed interface fluxes. These
properties have attracted the attention of the porous media flow
community over the past 15 years
\citep[see][]{riviere_2000,riviere_2002,bastian_2002} and are utilised
here.

In this paper, a novel CVFEM formulation which is conservative and
consistent is presented. The continuity equations are embedded into
the pressure equation to enforce mass conservation and the exact force
balance. The formulation employs an implicit algorithm with respect to
time that is less restrictive than the
implicit-pressure-explicit-saturation (IMPES) scheme often adopted in
porous media flow problems \citep[e.g. see][]{aziz_1986,
  geiger_2004}. There are two key novel aspects in our approach. The
first is that it uses a dual consistent CVFEM representation that is
embedded in the \PN[n]{m}[DG] family of triangular and tetrahedral
finite element pairs \citep[see][]{kaisu}. For this element type, the
velocity field is represented by $n^{\rm th}$-order polynomials that
are discontinuous across elements, while the pressure field is
represented by $m^{\rm th}$-order polynomials that may be continuous
(termed P$_n$DG-P$_m$) or discontinuous (termed P$_n$DG-P$_m$DG)
across elements. In particular, here the P$_n$DG-P$_{n+1}$ element
pair, which allows the velocity to exactly represent the pressure
gradients in the flow solution for homogeneous material properties, is
demonstrated. The \PNDG[n+1]{n} element pair, which has similar
properties to the \PN[n]{n+1} element pair, but allows a
representation in which pressure, saturation and other solution
variables are discontinuous across finite element boundaries, is also
demonstrated.  This solves the long-standing problem described above
with respect to the use of traditional CVFEM methods to capture sharp
changes in material properties. The second key novelty is the velocity
representation that is described in detail in
Section~\ref{overlapping_method_section} and can achieve an exact
representation of the multiphase Darcy force balance equations.

The remainder of this paper is organised as follows. The governing
equations and numerical methods employed to solve them are presented
in Section~\ref{overlapping_method_section}. Model set-up and results
including comparison against reference data for a series of benchmark
problems are given in Sections~\ref{setup} and \ref{res},
respectively. Finally, some concluding remarks are presented in
Section~\ref{conc}.

\section{Governing equations and Numerical Method}
\label{overlapping_method_section}

\subsection{A Novel Representation of Multiphase Darcy Flows}
Darcy's law for immiscible multiphase flow may be written in the form:
\begin{equation}\label{e:darcy_eqn}
  \mathbf{q}_{\alpha} =
  -\frac{\mathcal{K}_{{r}_\alpha}\mathbf{K}}{\mu_{\alpha}}\left(
  \nabla p_{\alpha} - {\mathbf{s}_{u}}_{\alpha} \right),
\end{equation}
where $\mathbf{q}_{\alpha}$ is the $\alpha$-th phase Darcy velocity,
$\mathbf{K}$ is the absolute permeability tensor of the porous medium,
$\mathcal{K}_{{r}_\alpha}\left(S_{\alpha}\right)$ is the phase
relative permeability, which is a function of the phase saturation
$S_{\alpha}\left(\mathbf{r},t\right)$. $\mu_{\alpha}$, $p_{\alpha}$
and $\mathbf{s}_{{u}_\alpha}$ are the phase dynamic viscosity,
pressure and source term, which may include gravity and capillary
pressure, respectively.

@@

In order to discretise Eqn.~\ref{force-bal}, a finite element
representation for $\mathbf{v}_\alpha$ and $p$ is assumed, expressed
in terms of their FE basis functions $Q_{j}$ and $P_{j}$,
respectively, as:
\begin{equation}
  \mathbf{v}_\alpha(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_u}
  Q_{j}(\bm{r})\mathbf{v}_{\alpha,j}(t) \;\;\;\;\text{ and } \;\;\;\;
  p(\bm{r},t) = \sum\limits_{j=1}^{\mathcal{N}_p}
  P_{j}(\bm{r})p_{j}(t).
\end{equation} 
Here $\mathcal{N}_{u}$ and $\mathcal{N}_{p}$ are the total number of
degrees of freedom for the FE force and pressure representations. Each
component of the weak form of the force balance (Eqn.~\ref{force-bal})
is tested with the $\mathbf{v}_\alpha$ basis function space to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_i \left (
             {\mathbf v}_\alpha + \nabla p -{\mathbf s}_{u_\alpha}
             \right) dV \right. + \displaystyle \oint_{\Gamma_{E}}
                   {Q}_i {\mathbf n} \left(p - \tilde{p}\right)
                   d\Gamma + \nonumber \\ \oint_{\Gamma_{\Omega}} {
                     Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma
                   = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of
element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of
the computational domain. The numerical pressure $\tilde{p}$ appearing
in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is
the arithmetic mean of the potentially discontinuous pressure across
the element $E$. This term vanishes when a continuous formulation is
used to discretise the pressure field. The last term in
Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure
level to $p_\text{bc}$ on a computational domain boundary.

@@

\subsection{Control-Volume Finite element types}
In the formulation presented here, pressure, velocity, permeability
and porosity are represented FE-wise, but saturation, relative
permeability and fluid properties such as viscosity and density are
represented CV-wise. Fig.~\ref{fem_cv_represent_a} displays the two
families of element types presented in this
paper. Fig.~\ref{fem_cv_represent_a} (a) shows the P$_{n}$DG-P$_{n+1}$
element type, with $n = 1$, in two dimensions (2D). Here the velocity
field is linear, discontinuous (between elements), the pressure field
has a quadratic, continuous (between elements), representation and the
CV's span various elements. Fig.~\ref{fem_cv_represent_a} (b) shows
the P$_{n+1}$DG-P$_{n}$DG element type, with $n = 1$, also in 2D. Here
the velocity field is quadratic, discontinuous (between elements), the
pressure field has a linear, discontinuous (between elements),
representation and the CV's do not span elements. Note that in the
latter element pair, CV's have piece-wise constant permeability and
porosity values.

\begin{figure}[h!]
\subfloat[P$_1$DG-P$_2$]{\includegraphics[width=.5\textwidth]{diagrams/p1dgp2-cont-sat}}
\subfloat[P$_2$DG-P$_1$DG]{\includegraphics[width=.5\textwidth]{diagrams/p2dgp1-dg-sat}}
\caption{2D representation of the element pairs presented in this
  work. Shaded areas denote control volumes (in which saturation is
  stored), black points represent the pressure nodes and white points
  the velocity nodes.}
    \label{fem_cv_represent_a}
\end{figure}

@@

\subsection{Saturation and Global Mass Conservative Equations} \label{Section:Saturation_Global}
Here, the saturation conservation equations are discretised and the
global mass balance equation is derived. In this work, fluids are
assumed incompressible. The saturation equation can be written as:
\begin{equation}
  \phi\displaystyle\frac{\partial S_{\alpha} }{\partial t} + \nabla
  \cdot \left( {\mathbf u}_{\alpha} S_{\alpha}\right) =
  s_{cty,\alpha},
  \label{saturation_equation}
\end{equation}
where $\phi$ is the porosity and $s_{cty}$ is a source
term. Eqn.~\ref{saturation_equation} is discretised in space by
testing with CV basis functions $M_{i}$ and with the $\theta$-method
in time:
\begin{eqnarray}
  \int_{\Omega} M_{i} \displaystyle\frac{\phi \left({S_{\alpha
        i}^{n+1}}-{S_{\alpha i}^{n}}\right)}{\Delta t} dV + \nonumber
  \\ \oint_{\Gamma_{CV_{i}}} \left[\theta^{n+\frac{1}{2}} {\mathbf n}
    \cdot {\mathbf u}_{\alpha}^{n+1} S_{\alpha}^{n+1} +
    \left(1-\theta^{n+\frac{1}{2}}\right) {\mathbf n} \cdot {\mathbf
      u}_{\alpha}^{n} S_{\alpha}^{n} \right]d\Gamma = \int_{\Omega}
  M_{i} {s_{cty,\alpha}^{n+\theta}} dV,
  \label{detail-sat-eqn-k}
\end{eqnarray}
where $\mathbf{n}$ is the outward-pointing unit normal vector to the
surface ($\Gamma_{CV_{i}}$) of the control volume $i$ and $n$ is the
current time level. More details on the numerical methods used to
solve Eqn.~\ref{detail-sat-eqn-k} can be found in
\citet{pavlidis_2013b}.

The global continuity equation is obtained by summing
Eqn.~\ref{detail-sat-eqn-k} over all phases:
\begin{eqnarray}
  \sum\limits_{\alpha=1}^{\mathcal{N}_{\alpha}} \left\lbrace
  \int_{\Omega} M_{i} \displaystyle\frac{\phi\left({S_{\alpha
        i}^{n+1}}-{S_{\alpha i}^{n}}\right) } {\Delta t} dV +
  \right. \nonumber \\ \left. \displaystyle\oint_{{\Gamma_{CV}}_{i}}
  \left[\theta^{n+\frac{1}{2}} {\mathbf n} \cdot {\mathbf
      u}_{\alpha}^{n+1} S_{\alpha}^{n+1} +
    \left(1-\theta^{n+\frac{1}{2}}\right) {\mathbf n} \cdot {\mathbf
      u}_{\alpha}^{n} S_{\alpha}^{n} \right] \,d\Gamma -
  \right. \nonumber \\ \left. \displaystyle\int_{\Omega} M_{i}
  s_{cty,\alpha}^{n+\theta}\, dV \right\rbrace = 0.
           \label{detail-sat-eqn-k-sum}
\end{eqnarray}
Eq.~\ref{detail-sat-eqn-k-sum} is also bounded by the constraint:
\begin{equation}
  \sum\limits_{\alpha=1}^{\mathcal{N}_{\alpha}} {S_{\alpha}^{n}}_{i} =
  1, \quad \forall n,
\end{equation}
Solving for ${\mathbf u}^{n+1}$ in matrix form,
Eqn.~\ref{detail-sat-eqn-k-sum} becomes:
\begin{equation}
  {\mathbf B}^T \underline{\bf v}^{n+1} = \underline{\bf s}_{p}.
  \label{glob-cty-matrix}
\end{equation}

\subsection{Velocity at Control Volume Interfaces} \label{opt-up} 
The velocity to be used at the interface between CV's in the
saturation conservation (Eqn.~\ref{detail-sat-eqn-k}) and global
continuity (Eqn.~\ref{detail-sat-eqn-k-sum}) equations needs to be
determined. On the CV faces, there is no information about the flow
direction as the velocity is discontinuous at the CV boundaries. The
discontinuities occur between (a) control volumes within each FE and
(b) between elements when using the P$_{n+1}$DG-P$_{n}$DG pair.
